---
title: "gov51-ps4"
author: "Joshua Stewart"
format: pdf
editor: visual
---

## Github: <https://github.com/JoshuaMcKoy/Gov51_PS4>

# 1 Setup and Exploration (10 points)

## 1.1 Load packages and data

**Load the tidyverse, estimatr, and modelsummary packages. Then load the dataset from
data/card1995.csv**

```{r}
library(tidyverse)
library(estimatr)
library(modelsummary)

card <- read_csv("data/raw/card1995.csv")
```

## 1.2 Describe the data

**How many observations are in the dataset? What is the unit of observation?**

```{r}
nrow(card)
```

There are 3,010 observations in the dataset. The unit of observation are individuals.

## 1.3 Distribution of years of education

**Create a histogram of educ. What is the mean? Are there any unusual spikes? What might explain them?**

```{r}
library(ggplot2)

ggplot(card, aes(x = educ)) +
  geom_histogram(binwidth = 1, fill = "pink", color = "black") +
  labs(title = "Distribution of Years of Education",
       x = "Years of Education",
       y = "Number of Respondents") +
  theme_minimal()

mean(card$educ, na.rm = TRUE)
```

The mean years of education in the sample is 13.26 years, with large spikes at 12 (high school diploma) and 16 years (undergraduate college diploma). I think this is because the data reflects major educational milestones in society.

## 1.4 The naïve relationship

**Create a scatter plot of lwage (y-axis) against educ (x-axis). Add a linear regression line.
Based on this plot, does more education appear to be associated with higher wages?**

```{r}
ggplot(card, aes(x = educ, y = lwage)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE, color = "pink") +
  labs(title = "Log Wages vs. Years of Education",
       x = "Years of Education",
       y = "Log Wage") +
  theme_minimal()
```

Given that log wage seems to increase as years of education increases, the scatter plot shows a positive relationship between years of education and log wages. Although there is some variation in wages at each level of education, the fitted regression line slopes upward, suggesting that individuals with more education tend to have higher wages.

## 1.5 Why OLS might be misleading

**The OLS regression of lwage on educ might not give us the causal effect of education. Name two specific reasons why people with more education might earn more even if education itself had no effect.**

People with more education might earn more even if education itself had no effect because of their inherent disposition (e.g., more "hard working" people may be more likely to earn higher wages with or without education, though their ability may motivate them to attain education).

Second, people from wealthier backgrounds may have more resources, allowing for them to stay in school longer, while providing them with better job opportunities and their family's network. This would suggest that this effect could be a product of familial advantages instead of a simple correlation between education and wage.

# 2 The Instrument: College Proximity (25 points)

## 2.1 What the instrument measures

**The variable nearc4 equals 1 if the respondent grew up near a 4-year college. What share of
respondents in this sample grew up near a 4-year college?**

```{r}
nrow(card %>% filter(nearc4 == 1))
nrow(card %>% filter(nearc4 == 1)) / nrow(card)
```

2,052 individuals, which is 68.2% of the respondents in this sample, grew up near a 4-year college.

## 2.2 Relevance: does proximity affect education?

**The first condition for a valid instrument is relevance: the instrument must be correlated with the endogenous variable (education). Run the following regression — the first stage — and report the coefficient on nearc4 along with its standard error and t-statistic.**

```{r}
first_stage <- lm_robust(
  educ ~ nearc4 + black + smsa + south + exper + expersq + married, 
  data = card, 
  se_type = "HC2")

summary(first_stage)
coef(first_stage)["nearc4"]
```

The coefficient for nearc4 is 0.327, which means that people who grew up near a 4-year college obtained approximately 0.33 more years of education on years, holding all variables constant. The standard error is 0.080, and the t-value is 4.061.

## 2.3 Checking instrument strength

**Report the t-statistic on nearc4 from the first-stage regression. Compute the first-stage F-statistic as 𝐹 = 𝑡2.**

**Is your instrument strong? (Rule of thumb: F \> 10 suggests a strong instrument.) Why does instrument strength matter for the reliability of the 2SLS estimate?**

```{r}
4.0607488^2
```

The first-stage F is approximately 16.49. Given that 16.49 is greater than 10, the instrument would be considered strong. Instrument strength matters for the reliability of the 2SLS estimate because if the instrument is weak, the 2SLS estimate could be biased.

## 2.4 Exclusion restriction

**The second condition for a valid instrument is the exclusion restriction: college proximity must affect wages only through its effect on education — not through any other channel**

**(a) State the exclusion restriction formally: what must be true about the relationship between nearc4 and lwage once we control for education?**

The exclusion restriction requires that, after controlling for education and other variables, nearc4 has no direct effect on wages. This means that proximity to a college affects wages only through its impact on education.

**b) Can you think of a way the exclusion restriction might be violated — that is, a reason why growing up near a college might affect your wages for reasons other than more education? Be specific.**

The exclusion restriction might be violated if growing up near a college is associated with better local labor market conditions, for some reason. For example, areas with colleges may have more job opportunities or higher wages because of union strikes, which may direclty increase earnings even without additional education.

## 2.5 Balance check

**If the instrument is as-good-as-randomly assigned, respondents who grew up near a college should look similar on pre-treatment characteristics to those who did not.**

**Compute the mean black, smsa, and south for the two groups (nearc4 == 0 and nearc4 == 1). Are the groups balanced? What does any imbalance tell you about the instrument?**

```{r}
card %>%
  group_by(nearc4) %>%
  summarize(
    mean_black = mean(black, na.rm = TRUE),
    mean_smsa  = mean(smsa, na.rm = TRUE),
    mean_south = mean(south, na.rm = TRUE))
```

The two groups are not balanced. The imbalance tells us that people who grew up near a college are more likely to live in urban areas and less likely to be from the South. These differences suggest that proximity to a college is correlated with observable characteristics, indicating that the instrument may not beas good-as-randomly assigned.

# 3 The Wald Estimator (15 points)

**The simplest IV estimate is the Wald estimator: the reduced-form effect divided by the
first-stage effect.**

## 3.1 Compute the reduced form

**Run the reduced form regression: regress lwage directly on nearc4 (without educ, but
including the same controls as the first stage). Report the coefficient on nearc4 and interpret
it: what does it tell us about how proximity to a college relates to wages?**

```{r}
model_reduc <- lm(lwage ~ nearc4 + black + smsa + south + exper + expersq + married, data = card)
summary(model_reduc)
coef(model_reduc)["nearc4"]
names(card)
```

The coefficient on nearc4 is 0.041, which indicates that individuals who grew up near a 4-year college earn about 4.1% higher wages, on average, holding other factors constant. That said, the p-value is quite low (0.0146), which demonstrates that the data is statistically significant. This suggests that proximity to college has a meaningful direct relationship with wages, and any effect is likely not operating through increased educational attainment rather than directly impacting earnings.

## 3.2 Calculate the Wald estimator by hand

**Using the first-stage coefficient from Section 2.2 and the reduced-form coefficient from Section
3.1, compute the Wald estimator by hand. Show your calculation.**

```{r}
coef(model_reduc)["nearc4"]/coef(first_stage)["nearc4"]
```

The Wald estimator is 0.0114, meaning that one additional year of education increases log wages by about 0.1249, or roughly 12.49%.

## 3.3 Interpret the Wald estimator

**Interpret your Wald estimate in a sentence. What does it say about the effect of one additional
year of education on wages?**

The Wald estimator means that one additional year of education increases log wages by about 0.125, or roughly 12.5%.

# 4 OLS vs. 2SLS (35 points)

## 4.1 OLS estimate

**Run OLS of lwage on educ and the controls. Report the coefficient on educ.**

```{r}
ols <- lm_robust(lwage ~ educ + black + smsa + south + exper + expersq + married,
                 data = card, 
                 se_type = "HC2")
summary(ols)
coef(ols)["educ"]
```

The coefficient of education in the OLS regression model is 0.071, which implies that one additional year of education is associated with approximately a 7.1% increase in wages, holding other factors constant.

## 4.2 2SLS estimate

**Run the 2SLS regression using iv_robust() from estimatr. The formula syntax separates
the structural equation (left of \|) from the first-stage instruments (right of \|):**

```{r}
iv <- iv_robust(lwage ~ educ + black + smsa + south + exper + expersq + married | nearc4 + black + smsa + south + exper + expersq + married,
                data = card, 
                se_type = "HC2")
summary(iv)
```

**Report the coefficient on educ. Does it match your Wald estimate from Section 3.2?**

The coefficient on education is 0.125, which matches my previous Wald estimate (0.125).

## 4.3 Side-by-side comparison

**Display both the OLS and 2SLS estimates in a single regression table using modelsummary().
Keep only the rows for educ and the model statistics (N, R²).**

```{r}
modelsummary(list("OLS" = ols, "2SLS" = iv),
             coef_map = c("educ" = "Years of Education"),
             gof_map = c("nobs", "r.squared"))
```

## 4.4 Why do OLS and 2SLS differ?

**The OLS and 2SLS estimates of the return to education are likely to differ. Answer the
following:**

**(a) Which is larger — OLS or 2SLS?**

The 2SLS estimate is larger.

**(b) Card (1995) interprets the finding that 2SLS \> OLS as evidence of downward bias in OLS. What could cause OLS to underestimate the return to education? (Hint: think about measurement error in educ, or about who the compliers are.)**

One reason OLS may underestimate the return to education is measurement error in education, which could bias the OLS estimate toward zero. Additionally, the IV estimate identifies the effect for compliers, who may inherently experience larger returns to education than the average individual, making the 2SLS estimate larger than OLS.

**(c) Which estimate is more credible as a causal estimate of the effect of education on wages?
Explain briefly.**

The 2SLS estimate is more credible as a causal estimate because it uses proximity to college as an instrument for education, helping to address omitted variable bias and measurement error. In contrast, the OLS estimate may be biased due to unobserved factors like ability or family background.

## 4.5 Adding IQ as a control

**The dataset includes IQ scores for a subset of respondents.**

**(a) Run OLS of lwage on educ, IQ, and all other controls. How does including IQ change the OLS coefficient on educ?**

```{r}
ols_iq <- lm_robust(lwage ~ educ + IQ + black + smsa + south + exper + expersq + married,
                    data = card,
                    se_type = "HC2")

summary(ols_iq)
coef(ols_iq)["educ"]
```

Including the IQ lowers the OLS coefficeint on education, which makes sense given that the IQ variable absorbs some fo the explanatory power. The new education coefficient is 0.0676, which suggests that one additional year of education is associated with approximately a 6.76% increase in wages, holding other factors constant.

**(b) Now run 2SLS including IQ. Does the 2SLS estimate change substantially when IQ is added?**

```{r}
SLS_iq <- iv_robust(lwage ~ educ + IQ + black + smsa + south + exper + expersq + married | nearc4 + IQ + black + smsa + south + exper + expersq + married,
                data = card, 
                se_type = "HC2")
summary(SLS_iq)
coef(iv)["educ"]
coef(SLS_iq)["educ"]
```

The 2SLS estimate including IQ decreases the coefficicent of educ froom 0.1249 to 0.1094.

**(c) What does the comparison between part (a) and part (b) suggest about the source of the discrepancy between OLS and IV?**

Including IQ slightly reduces both the OLS and 2SLS estimates of the return to education. However, the gap between the OLS and IV estimates remains large even after controlling for IQ. This suggests that differences in ability, as measured by IQ, are not the primary source of the discrepancy between OLS and IV.

# 5 Interpretation and LATE (15 points)

## 5.1 Who are the compliers?

**In a randomized experiment, the ATE averages over everyone. But IV estimates a Local Average Treatment Effect (LATE) — the average effect for compliers: units whose treatment status changes when the instrument changes.**

**In this context, who are the compliers? Describe them in concrete terms: what type of
person’s educational attainment was increased because they happened to grow up near a 4-
year college?**

In this context, the compliers are people who decision to attend college is infuenced by their proximity to a college. For example, they are the type of people who would not have attended college if they had not grown up next to one, but who do attend college because of their proximity.

## 5.2 Write the LATE interpretation

**Write a single careful sentence interpreting the 2SLS coefficient as a LATE. Your sentence
should specify:
(i) The causal effect being measured
(ii) For which specific group of people
(iii) What source of variation identifies it**

The 2SLS coefficient estimates the causal effect of an additional year of education on wages for individuals whose decision to attend school are influenced by their proximity to a 4-year college, with identification coming from variation in education generated by growing up near versus far from a college.

## 5.3 External validity

**Card finds a 2SLS estimate around 12–13% return to each additional year of education. Is this LATE likely to be the same as the ATE (the return for the average person in the sample)?**

**Think about whether the compliers — men who went to more school because they happened to live near a college — are likely to have higher or lower returns to education than the typical person. Explain your reasoning.**

My assumption is that this LATE is unlikely to be the same as the ATE, given that the LATE only applies to compliers rather than the entire population. Compliers are unqiue in the sense that they are likely on the margin of attending college and may encounter higher costs or constraints, so the additional education they receive could generate relatively larger returns compared to the average person. Thus, the 2SLS estimate may overstate the return to education for the broader population.
